Computational methods in landscape ecology

### 2024-05-22 Manuscript.qmd
Minor changes intro

---
title: "Computational methods in landscape ecology"
keywords: keyword1, keyword2
author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg (Austria)
format:
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header: 
      text: | 
        \usepackage{lineno}\linenumbers
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
papersize: a4paper
fontsize: 12pt
geometry: margin=25mm
indent: 2m
linestretch: 1.75
bibliography: CLER.bib
csl: springer-basic-author-date.csl
---

>>>>

---
title: "Computational methods in landscape ecology"
keywords: keyword1, keyword2
author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg (Austria)
format:
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header: 
      text: | 
        \usepackage{lineno}\linenumbers
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
papersize: a4paper
fontsize: 12pt
geometry: margin=25mm
indent: 2m
linestretch: 1.5
bibliography: CLER.bib
csl: springer-basic-author-date.csl
---

### 2024-05-22 Manuscript.tex
Minor changes intro

\setstretch{1.75}
\section{Introduction}\label{introduction}

Landscapes are typically defined as mosaics of different land covers,
habitats, ecosystems, or land use systems (Forman and Godron 1986;
Forman 1995) - with empathize on existing heterogeneity in at least one
factor of interest (Turner and Gardner 2015). Linking the spatial
heterogeneity and abiotic as well as biotic processes is the fundamental
concept of landscape ecology. This includes potential feedbacks between
heterogeneity and processes, spatial and temporal flows across
heterogenous landscapes, or the management of heterogeneity in
landscapes (Risser et al. 1984; Turner 1989). However, due to the
spatial context of heterogeneity, the pattern-process link is scale
dependent (Wiens 1989; Levin 1992; Wu 2013).

Computational science involves analyzing abstracted core mechanisms of
research questions using data and algorithms. Thus, computational
ecology can be defined if computational science is used to address
ecological research questions with focus on complex adaptive systems and
data-driven approaches (Poisot et al. 2019). Computational ecology in
general and simulations specifically are important for ecology because
ecological data is often context- and scale-dependent and infeasible to
study in reproducible, replicable and controlled experimental settings
(Petrovskii and Petrovskaya 2012).

Because landscape ecology is a cross-disciplinary field including (e.g.,
social sciences, geography, or ecology and evolution Wiens 1997; With
2019), the general increase of available data (Chi et al. 2016; Jarić et
al. 2020; Wüest et al. 2020; Nathan et al. 2022), or the complex systems
characteristics of landscapes and processes (Newman et al. 2019), there
is the need for sophisticated computational methods in order to link
patterns and processes.

This paper aims to introduce the latest developments of computation
methods in landscape ecology. However, it is not a general introduction
into (computational) landscape ecology. For a more comprehensive
introductions to landscape ecology in general, please see e.g.~Turner
and Gardner (2015) or Gergel and Turner (2017).

@@

\bibitem[\citeproctext]{ref-Levin1992}
Levin SA (1992) The problem of pattern and scale in ecology. Ecology
73:1943--1967. \url{https://doi.org/10.2307/1941447}

@@

@@

@@

\bibitem[\citeproctext]{ref-Wiens1989}
Wiens JA (1989) Spatial scaling in ecology. Functional Ecology
3:385--397. \url{https://doi.org/10.2307/2389612}

@@

@@

>>>>

\setstretch{1.5}
\section{Introduction}\label{introduction}

Landscapes are typically defined as mosaics of different land covers,
habitats, ecosystems, or land use systems (Forman and Godron 1986;
Forman 1995) - with empathize on heterogeneity of at least one factor of
interest (Turner and Gardner 2015). Linking the spatial heterogeneity
and ecological processes, including potential feedbacks, is the
fundamental concept of landscape ecology (Risser et al. 1984; Turner
1989, 2005). Typical research topics include ecological flows in
landscape mosaics, drivers and consequences of land use and land cover
change, nonlinear dynamics and landscape complexity, scale and scaling,
spatial analysis and landscape modeling, linking landscape metrics to
ecological processes, integrating humans and their activities into
landscape ecology, landscape conservation and sustainability, or
accuracy assessment and uncertainty analysis (Wu and Hobbs 2002; Wu
2013).

Computational science involves analyzing abstracted core mechanisms of
research questions using data and algorithms. Thus, computational
ecology can be defined as computational science that is used to address
ecological research questions with focus on complex, adaptive systems
and data-driven approaches (Poisot et al. 2019). Computational ecology
in general and simulations specifically are of importance for ecology
because ecological data is often context- and scale-dependent and
unfeasible to study in reproducible, replicable, and controllable
experimental settings (Petrovskii and Petrovskaya 2012; but see Wiersma
2022). Because landscape ecology is a cross-disciplinary field
including, e.g., social sciences, geography, or ecology and evolution
(Wiens 1997; With 2019), the general increase of available data (Chi et
al. 2016; Jarić et al. 2020; Wüest et al. 2020; Nathan et al. 2022), or
the complex landscape systems (Newman et al. 2019), there is the need
for sophisticated computational methods in order to link patterns and
processes. Thus, computational methods are one of the most important
tools of modern scientific research (Prlić and Procter 2012; Wilson et
al. 2014).

This paper aims to introduce the latest developments of computation
methods in landscape ecology. However, it is neither a systematic review
of the literature nor general introduction to (computational) landscape
ecology. For a more comprehensive introductions to landscape ecology in
general, please see Turner and Gardner (2015) or Gergel and Turner
(2017).

@@

@@

\bibitem[\citeproctext]{ref-Prlic2012}
Prlić A, Procter JB (2012) Ten simple rules for the open development of
scientific software. PLoS Computational Biology 8:e1002802.
\url{https://doi.org/10.1371/journal.pcbi.1002802}

@@

\bibitem[\citeproctext]{ref-Turner2005}
Turner MG (2005) Landscape ecology: What is the state of the science?
Annual Review of Ecology, Evolution, and Systematics 36:319--344.
\url{https://doi.org/10.1146/annurev.ecolsys.36.102003.152614}

@@

@@

\bibitem[\citeproctext]{ref-Wiersma2022}
Wiersma YF (2022) A review of landscape ecology experiments to
understand ecological processes. Ecological Processes 11:57.
\url{https://doi.org/10.1186/s13717-022-00401-0}

\bibitem[\citeproctext]{ref-Wilson2014}
Wilson G, Aruliah DA, Brown CT, et al (2014) Best practices for
scientific computing. PLoS Biology 12:e1001745.
\url{https://doi.org/10.1371/journal.pbio.1001745}

@@

\bibitem[\citeproctext]{ref-Wu2002}
Wu J, Hobbs R (2002) Key issues and research priorities in landscape
ecology: An idiosyncratic synthesis. Landscape Ecology 17:355--365.
\url{https://doi.org/10.1023/A:1020561630963}

### 2024-05-22 Sections/introduction.qmd
Minor changes intro

Landscape ecology is fun [@Turner1989].
There are many things to explore [@With2019].

>>>>

Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986; @Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
Linking the spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989; @Turner2005].
Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002; @Wu2013].

<!-- MH: Currently using "computational science", "computational ecology", and "computational methods". Streamline terms?--> 
Computational science involves analyzing abstracted core mechanisms of research questions using data and algorithms.
Thus, computational ecology can be defined as computational science that is used to address ecological research questions with focus on complex, adaptive systems and data-driven approaches [@Poisot2019].
Computational ecology in general and simulations specifically are of importance for ecology because ecological data is often context- and scale-dependent and unfeasible to study in reproducible, replicable, and controllable experimental settings [@Petrovskii2012; but see @Wiersma2022]. Because landscape ecology is a cross-disciplinary field including, e.g., social sciences, geography, or ecology and evolution [@Wiens1997; @With2019], the general increase of available data [@Chi2016; @Jaric2020; @Nathan2022; @Wuest2020], or the complex landscape systems [@Newman2019], there is the need for sophisticated computational methods in order to link patterns and processes. Thus, computational methods are one of the most important tools of modern scientific research [@Prlic2012; @Wilson2014].

This paper aims to introduce the latest developments of computation methods in landscape ecology.
However, it is neither a systematic review of the literature nor general introduction to (computational) landscape ecology.
For a more comprehensive introductions to landscape ecology in general, please see Turner and Gardner [-@Turner2015] or Gergel and Turner [-@Gergel2017].
